---
title: "Exploratory Data Analysis"
author: "Cass Crews"
format: html
editor: visual
---

# Introduction

# Loading Packages and Reading in Data

```{r}
library(tidyverse)
library(grid)
library(gridExtra)
```

```{r}
raw_data<-read_csv("diabetes_binary_health_indicators_BRFSS2015.csv") |>
  select(Diabetes_binary, HighBP, HighChol, PhysActivity, Fruits, Veggies, Smoker, HvyAlcoholConsump, Sex, Age, BMI, Income)
```



# Initial Cleaning

We need to complete a few data validation steps to determine whether the data need to be cleaned before any summaries and analysis. To start, we will count the number of missing values for each variable. 

```{r}
raw_data |>
  summarize(across(everything(),~sum(is.na(.x))))
```

Fortunately, there are no missing values, so we can proceed to confirming each variable's data type is appropriate. 

```{r}
str(raw_data)
```

All of these variables except BMI should actually be factors, but they are stored as numeric variables. Let's convert them to factors and give them informative labels. However, let's first confirm the levels of these factors are what we expect. 

```{r}
apply(raw_data |> select(!BMI), MARGIN = 2, unique)
```

These values are exactly as we expect: the first nine variables are binary. The `Age` variable is a multi-level factor ranging from a value of `1` for individuals between 18 and 24 to a value of `13` for those at least 80 years old. The `Income` variable is also a multi-level factor ranging from a value of `1` for individuals with household incomes below \$10,000 to a value of `8` for individuals with household incomes above \$75,000. 

Now we need to ensure the `BMI` variable takes on reasonable values, which we can do by extracting the minimum and maximum within our sample. 

```{r}
raw_data |>
  summarize(across(BMI, list(min = min, max = max), .names = "{.col}_{.fn}"))
```

While a BMI of 98 is very high, it is still plausible. Thus, we can reasonably assume the BMI values are correct. 

Thankfully, these data are very clean. We are now ready to convert most of the variables to factors. In doing so, let's give the variables more consistent names. 

```{r}
data<-raw_data |>
  mutate(diabetes_binary = factor(Diabetes_binary, levels = 0:1, labels = c("No", "Yes"))) |>
  mutate(high_bp = factor(HighBP, levels = 0:1, labels = c("No", "Yes"))) |>
  mutate(high_chol = factor(HighChol, levels = 0:1, labels = c("No", "Yes"))) |>
  mutate(phys_activity = factor(PhysActivity, levels = 0:1, labels = c("No", "Yes"))) |>
  mutate(fruits = factor(Fruits, levels = 0:1, labels = c("No", "Yes"))) |>
  mutate(veggies = factor(Veggies, levels = 0:1, labels = c("No", "Yes"))) |>
  mutate(smoker = factor(Smoker, levels = 0:1, labels = c("No", "Yes"))) |>
  mutate(hvy_alc_consump = factor(HvyAlcoholConsump, levels = 0:1, labels = c("No", "Yes"))) |>
  mutate(sex = factor(Sex, levels = 0:1, labels = c("Female", "Male"))) |>
  mutate(age = factor(Age, levels = 1:13, labels = c("18 to 24", "25 to 29",
                                                     "30 to 34", "35 to 39",
                                                     "40 to 44", "45 to 49",
                                                     "50 to 54", "55 to 59",
                                                     "60 to 64", "65 to 69",
                                                     "70 to 74", "75 to 79",
                                                     "80+"))) |> 
  mutate(income = factor(Income, levels = 1:8, labels = c("Less than $10k", "$10k to $15k",
                                                     "$15k to $20k", "$20k to $25k",
                                                     "$25k to $35k", "$35k to $50k",
                                                     "$50k to $75k", "$75k+"))) |> 
  rename(bmi = BMI) |>
  select(!c(Diabetes_binary:Age, Income))

str(data)
```
Note that when we printed the data structure this time, each variable except `bmi` is a factor with interpretable labels. 

# Summaries

### Univariate Summaries

We are now ready to summarize the data. Let's start by exploring the unconditional distribution of each variable. 

For our binary categorical variables other than `sex`, let's produce a single table to compare the breakdown across `yes` and `no`. 

```{r}
data |>
  select(!c(age, sex, bmi, income)) |>
  pivot_longer(cols = everything(), names_to = "variables", values_to = "values") |>
  group_by(variables, values) |>
  summarize(count = n()) |>
  pivot_wider(names_from = values, values_from = count)
```

Looking at these univariate distributions, one key finding is that we have substantial class imbalance in our response variable. However, this is not particularly concerning given the large sample size. 

Before we move to multi-variate distributions, let's look at the univariate breakdowns for `sex`, `age`, and `income`, as well as some numeric summaries for `bmi`. 

```{r}
data |>
  group_by(sex) |>
  summarize(count = n()) 
```

It is notable that 56% of the observations correspond to females, compared to roughly 51% of the national population. 

```{r}
data |>
  group_by(age) |>
  summarize(count = n()) |>
  print(n = 13)
```

Given that these data are built on medical records, observation counts generally trend up with age until we reach the age groups near the average life expectancy. 

```{r}
data |>
  group_by(income) |>
  summarize(count = n())
```

For household income, we see that the most common income group is \$75,000+. Thus, these income groupings may not be particularly helpful for segmenting the population our sample represents, as we may be missing nuanced differences across income intervals above $75,000. 

While we have already captured the minimum and maximum BMI values, we should explore a few other summary statistics for our lone numeric variable. 

```{r}
data |>
  summarize(across(bmi, list(mean = mean, median =median, sd = sd, IQR = IQR)))
```

Perhaps surprisingly, given that BMI is naturally bounded at the low end, the BMI values in our sample are not excessively skewed; the mean and median are relatively close together. 

### Multivariate Summaries

To start our multivariate analysis, let's plot our response by each level of our `yes`-`no` variables. Given that we have already explored the unconditional distributions for these variables, we will use filled bar plots to focus specifically on the differences in diabetes rates across these conditional distributions.  

```{r fig.height=8}
g1<-ggplot(data = data, aes(x = phys_activity, fill = diabetes_binary)) + geom_bar(position = "fill") + labs(x = "Physical Activity?", y = "Share of Obs.", fill = "Diabetes?")

g2<-ggplot(data = data, aes(x = high_bp, fill = diabetes_binary)) + geom_bar(position = "fill") + labs(x = "High Blood Pressure?", y = "Share of Obs.", fill = "Diabetes?")

g3<-ggplot(data = data, aes(x = high_chol, fill = diabetes_binary)) + geom_bar(position = "fill") + labs(x = "High Cholesterol?", y = "Share of Obs.", fill = "Diabetes?")

g4<-ggplot(data = data, aes(x = fruits, fill = diabetes_binary)) + geom_bar(position = "fill") + labs(x = "Eat Fruit?", y = "Share of Obs.", fill = "Diabetes?")

g5<-ggplot(data = data, aes(x = veggies, fill = diabetes_binary)) + geom_bar(position = "fill") + labs(x = "Eat Veggies?", y = "Share of Obs.", fill = "Diabetes?")

g6<-ggplot(data = data, aes(x = smoker, fill = diabetes_binary)) + geom_bar(position = "fill") + labs(x = "Smoking History?", y = "Share of Obs.", fill = "Diabetes?")

g7<-ggplot(data = data, aes(x = hvy_alc_consump, fill = diabetes_binary)) + geom_bar(position = "fill") + labs(x = "Heavy Alcohol Consumption?", y = "Share of Obs.", fill = "Diabetes?")

grid.arrange(g1, g2, g3, g4, g5, g6, g7, ncol = 2, top = textGrob("Diabetes Frequency across Levels of Behavioral Variables", hjust = 0.59, gp = gpar(fontsize = 14)))
```

It seems that, unconditional on other variables, physical activity and heavy alcohol consumption are the behaviors most closely related to changes in diabetes frequency. Perhaps surprisingly, the diabetes rate is lower among individuals with relatively high alcohol consumption. 

Diabetes rates are also much higher among those with high cholesterol and high blood pressure. These must be common co-morbidities for diabetes. 

Let's now explore any differences in diabetes frequency across sex to see whether this demographic characteristic explains diabetes incidence on its own. 

```{r}
ggplot(data = data, aes(x = sex, fill = diabetes_binary)) + geom_bar(position = "fill") + labs(x = "Sex", y = "Share of Obs.", fill = "Diabetes?", title = "Diabetes Frequency by Sex")
```

It seems males in this dataset are slighly more likely to have diabetes. 

Now let's compare diabetes frequencies across age groups. 

```{r}
ggplot(data = data, aes(x = age, fill = diabetes_binary)) + geom_bar(position = "fill") + labs(x = "Age Group", y = "Share of Observations", fill = "Diabetes?", title = "Diabetes Frequency by Age Group") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

At least in our sample, diabetes rates increase with age until we reach the oldest age groups; the decline for older ages may reflect reduced life expectancy due to diabetes. 

Could household income be a predictor of diabetes? If so, there could be important implications for health interventions. 

```{r}
ggplot(data = data, aes(x = income, fill = diabetes_binary)) + geom_bar(position = "fill") + labs(x = "Income Bracket", y = "Share of Observations", fill = "Diabetes?", title = "Diabetes Frequency by Household Income Bracket") + 
  theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1))
```

It seems diabetes risk declines as household income rises. This may indicate that access to less processed foods, which can be more expensive, reduces the risk of diabetes. However, it could also be the case the higher incomes correspond with increased access to early detection tools and preventative treatments. 

For our final visuals summary, let's produce BMI box plots for those with and without diabetes to see if the BMI distributions vary. 

```{r}
ggplot(data = data, aes(x = diabetes_binary, y = bmi, fill = diabetes_binary)) + geom_boxplot() + 
  labs(title = "BMI Distribution by Diabetes Diagnosis", y = "BMI", x = "Diabetes?") + theme(legend.position = "none")
```

It does seem that those with diabetes have higher BMIs on average, but these distributions almost completely overlap. 

# Wrap-Up

This concludes our exploratory data analysis. We will now attempt to predict whether or not an individual has diabetes using only the variables explored on this page. 