---
title: "Exploratory Data Analysis"
author: "Cass Crews"
format: html
editor: visual
---

# Introduction

# Loading Packages and Reading in Data

```{r}
library(tidyverse)
```

```{r}
raw_data<-read_csv("diabetes_binary_health_indicators_BRFSS2015.csv") |>
  select(Diabetes_binary, PhysActivity, Fruits, Veggies, Smoker, HvyAlcoholConsump, Sex, Age)
```



# Initial Cleaning

We need to complete a few data validation steps to determine whether the data need to be cleaned before any summaries and analysis. To start, we will count the number of missing values for each variable. 

```{r}
raw_data |>
  summarize(across(everything(),~sum(is.na(.x))))
```

Fortunately, there are no missing values, so we can proceed to confirming each variable's data type is appropriate. 

```{r}
str(raw_data)
```

All of these variables should actually be factors, but they are stored as numeric variables. Let's convert them to factors and give them informative labels. However, let's first confirm the levels of these factors are what we expect. 

```{r}
apply(raw_data, MARGIN = 2, unique)
```

These values are exactly as we expect: the first seven variables are binary, while the final variable (`Age`) is a multi-level factor ranging from a value of `1` for individuals between 18 and 24 to a value of `13` for those at least 80 years old. 

Thankfully, these data are very clean. We are now ready to convert these variables to factors. In doing so, let's give the variables more consistent names. 

```{r}
data<-raw_data |>
  mutate(diabetes_binary = factor(Diabetes_binary, levels = 0:1, labels = c("No", "Yes"))) |>
  mutate(phys_activity = factor(PhysActivity, levels = 0:1, labels = c("No", "Yes"))) |>
  mutate(fruits = factor(Fruits, levels = 0:1, labels = c("No", "Yes"))) |>
  mutate(veggies = factor(Veggies, levels = 0:1, labels = c("No", "Yes"))) |>
  mutate(smoker = factor(Smoker, levels = 0:1, labels = c("No", "Yes"))) |>
  mutate(hvy_alc_consump = factor(HvyAlcoholConsump, levels = 0:1, labels = c("No", "Yes"))) |>
  mutate(sex = factor(Sex, levels = 0:1, labels = c("Female", "Male"))) |>
  mutate(age = factor(Age, levels = 1:13, labels = c("18 to 24", "25 to 29",
                                                     "30 to 34", "35 to 39",
                                                     "40 to 44", "45 to 49",
                                                     "50 to 54", "55 to 59",
                                                     "60 to 64", "65 to 69",
                                                     "70 to 74", "75 to 79",
                                                     "80+"))) |>
  select(!(Diabetes_binary:Age))

str(data)
```
Note that when we printed the data structure this time, each variable is a factor with interpretable labels. 

# Summaries

We are now ready to summarize the data. Let's start by exploring the unconditional distribution of each variable. 

For our binary categorical variables other than `sex`, let's produce a single table to compare the breakdown across `yes` and `no`. 

```{r}
data |>
  select(!c(age, sex)) |>
  pivot_longer(cols = everything(), names_to = "variables", values_to = "values") |>
  group_by(variables, values) |>
  summarize(count = n()) |>
  pivot_wider(names_from = values, values_from = count)
```

Looking at these univariate distributions, one key finding is that we have substantial class imbalance in our response variable. However, this is not particularly concerning given the large sample size. 

Before we move to multi-variate distributions, let's look at the univariate breakdowns for `sex` and `age`. 

```{r}
data |>
  group_by(sex) |>
  summarize(count = n()) 
```

It is notable that 56% of the observations correspond to females, compared to roughly 51% of the national population. 

```{r}
data |>
  group_by(age) |>
  summarize(count = n()) |>
  print(n = 13)
```

Given that these data are built on medical records, observation counts generally trend up with age until we reach the age groups near the average life expectancy. 

